---
title: "Getting started with ggplot2"
author: "Einar Hjörleifsson"
date: "29 January 2016"
output: 
  slidy_presentation:
    fig_height: 3
    fig_width: 4
---

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(tidyr) # fix to get rid of rude comments when loading dplyr
```

## Preamble

### Needed libraries for this tutorial:
```{r, message = FALSE, warning = FALSE, results='hide'}
library(geo)
library(ggplot2)
library(dplyr)
```

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
source("R/common.R")
```

### Getting example data into R
```{r}
d <- read.csv("http://www.hafro.is/~einarhj/data/minke.csv",
              stringsAsFactors = FALSE) %>% 
  tbl_df()
d2 <- read.csv("http://data.hafro.is/assmt/2015/cod/summary.csv") %>% 
  tbl_df()
```

#### The minke data

```{r}
str(d)
```

#### The assessment summary data

Taken from <a href="http://data.hafro.is/assmt/2015/cod/" target="_blank">hafro data site</a>

```{r}
str(d2)
```

## ggplot2 book - introduction

[Introduction](https://github.com/hadley/ggplot2-book/blob/master/introduction.rmd)

## Key components

"Every ggplot2 plot has three key components: 

1.  __data__,

2. A set of __aesthetic mappings__ [aes] between variables in the data and 
   visual properties, and 

3. At least one layer which describes how to render each observation. Layers
   are usually created with a __geom__ function."
`r columns(1)`
```{r}
ggplot(d, aes(x = age, y = length)) + 
  geom_point()
```

Different syntax, equivalent outcome:
```{r, eval = FALSE}
ggplot(d, aes(x = age, y = length)) + geom_point()
ggplot(d, aes(age, length))         + geom_point()
ggplot()                            + geom_point(data = d, aes(age, length))
ggplot(data = d)                    + geom_point(aes(x = age, y = length))
ggplot(d)                           + geom_point(aes(age, length))
d %>% ggplot()                      + geom_point(aes(age, length))
```

Can be stored as an object for later use:
```{r}
p <- ggplot(d, aes(age, length)) + geom_point()
```

The class and the structure:
```{r}
class(p)
str(p)
```

## aesthetic: colour
`r columns(2)`
```{r}
p <- ggplot(d)
p + geom_point(aes(age, length, colour = sex))
p + geom_point(aes(age, length, colour = area))
```

Manual control of colours:
```{r}
p <- ggplot(d)
p + geom_point(aes(age, length, colour = sex)) +
  scale_colour_manual(values = c("orange","brown"))
p + geom_point(aes(age, length, colour = area)) +
  scale_colour_manual(values = c("orange","brown"))
```

## aesthetic: shape
`r columns(1)`
```{r}
p + geom_point(aes(age, length, shape = sex))
```

## aesthetic: size
`r columns(1)`
```{r}
p + geom_point(aes(age, length, size = stomach.volume))
```

To reveal overlays:
```{r}
p + geom_point(aes(age, length, size = stomach.volume), alpha = 0.6)
```

## facetting

Display same graph by subsets of the data:
`r columns(1)`
```{r}
ggplot(d) + 
  geom_point(aes(age, length, colour = sex)) + 
  facet_wrap(~ area)
```

### Exercise

Create a code that results in this plot:
`r columns(1)`
```{r, echo = FALSE}
ggplot(d) +
  geom_point(aes(age, length, colour = area)) + 
  facet_wrap(~ sex)
```

## layer: add smoother to a plot

`r columns(2)`
```{r}
p <- ggplot(d, aes(age, length))
p + geom_point() + geom_smooth()
p + geom_point() + geom_smooth(span = 0.1)
```

Specific models:
`r columns(1)`
```{r}
p + geom_point() + geom_smooth(method = "lm")
p + geom_point() + geom_smooth(method = "lm")
```

## layer: points as jitters

Add a little random noise to the data to avoid overplotting:
`r columns(2)`
```{r}
p <- ggplot(d, aes(sex, length))
p + geom_point()
p + geom_jitter()
```

## layer: summarise via boxplots or violins
`r columns(2)`
```{r}
p + geom_boxplot()
p + geom_violin()
```

## layer: overlay data and summary layers

`r columns(1)`
```{r}
p + geom_violin() + geom_boxplot() + geom_jitter()
```

## layer: frequency polygons

`r columns(2)`
```{r}
p <- ggplot(d, aes(length))
p + geom_freqpoly()
p + geom_freqpoly(aes(colour = sex))
```

## layer: histograms

More details distribution than boxplots do, but more space:
`r columns(2)`
```{r}
p + geom_histogram()
p + geom_histogram(aes(fill = sex))
```

Better way for histograms:
`r columns(1)`
```{r}
p + geom_histogram(aes(fill = sex)) + facet_wrap(~ sex, ncol = 1)
```

Controlling binwidths:
```{r}
p + geom_histogram(binwidth = 50)
```

## geom: bar charts

Histograms for discrete data:
`r columns(1)`
```{r}
ggplot(d, aes(maturity)) + geom_bar()
```

## layer: line
`r columns(1)`
```{r}
ggplot(d2, aes(Year, Fbar)) + geom_line()
```

## layer: path
`r columns(2)`
```{r}
ggplot(d2, aes(SSB, Fbar)) + geom_line()
ggplot(d2, aes(SSB, Fbar)) + geom_path()
```

Check out difference of:
```{r, eval = FALSE}
?geom_line
?geom_path
```

Of more use:
`r columns(2)`
```{r}
ggplot(d2, aes(SSB, Fbar)) + geom_path() + geom_point(aes(colour = Year), size = 3)
```


## axes: labels

`r columns(2)`
```{r}
p <- ggplot(d, aes(age, length, colour = sex)) + geom_point()
p
p + labs(x = "Age [year]", y = "Length [cm]", colour = "Sex")
```

`r columns(2)`
```{r}
p + labs(x = NULL, y = NULL, colour = NULL)
p + labs(title = "My minke plot")
```

## axes: breaks

Controls which values appear as tick marks
`r columns(3)`
```{r}
p <- ggplot(d, aes(age, length)) + geom_point() + labs(x = NULL, y = NULL)
p
p +
  scale_x_continuous(breaks = c(5, 10, 15, 20, 25, 30, 35, 40, 45))
p +
  scale_x_continuous(breaks = seq(5, 45, by = 5)) +
  scale_y_continuous(breaks = seq(500, 950, by = 50))
```


## axes: limits

`r columns(3)`
```{r}
p <- ggplot(d, aes(maturity, length))
p + geom_jitter()
p + geom_jitter() + ylim(600, 800)
p + geom_jitter() + ylim(NA, 800) # setting only one limit
```

For descrete variables:
`r columns(1)`
```{r}
p + geom_jitter() + ylim(600,800) + xlim("immature","mature")
```

## axes: limits - warning

But be careful when using with summary statistics, e.g.:
`r columns(2)`
```{r}
p + geom_boxplot()
p + geom_boxplot() + ylim(600, 800)
```

Remedy:
`r columns(1)`
```{r}
p + geom_boxplot() + coord_cartesian(ylim = c(600, 800))
```

## A little gis

`r columns(2)`
```{r}
p <- ggplot(island, aes(lon, lat)) + labs(x = NULL, y = NULL)
p + geom_point()
p + geom_line()
```

```{r}
p + geom_path()
p + geom_path() + coord_map()
```

Base map:
`r columns(1)`
```{r}
m <- ggplot(island, aes(lon, lat)) +
  theme_bw() +
  geom_polygon(fill = "grey90") +
  coord_map() +
  labs(x = NULL, y = NULL)
m
```

## gis:
`r columns(2)`
```{r}
m + geom_point(data = d, aes(lon, lat))
m + geom_point(data = d, aes(lon, lat, colour = area))
```

```{r}
m + geom_point(data = d, aes(lon, lat, colour = sex))
m + geom_point(data = d, aes(lon, lat, colour = year))
m + geom_point(data = d, aes(lon, lat, colour = factor(year)))
m + geom_point(data = d, aes(lon, lat, size = length), alpha = 0.2)
# remedy of the above plot - is beyound the basic introduction
m + geom_point(data = d, aes(lon, lat, colour = length)) +
  scale_colour_gradient(low = "yellow", high = "red")
```

Another type of base map:
```{r}
library(ggmap)
m2 <- get_map(location = c(-19,65), zoom= 6, maptype = "satellite", source = "google")
m2 <- ggmap(m2) +
  labs(x = NULL, y = NULL)
m2
m2 + geom_point(data = d, aes(lon, lat), colour = "red")
```


## Size
`r columns(2)`
```{r}
p <- ggplot(d)
p + geom_point(aes(age, length))
p + geom_point(aes(age, length), size = 4)
```

## Colours
```{r}
p + geom_point(aes(age, length), colour = "blue")
# note the difference
p + geom_point(aes(age, length, colour = "blue"))
```
