---
title: "Hafró goes R (remix)"
author: "Bjarki Þór Elvarsson and Einar Hjörleifsson"
date: "January 19, 2016"
output:
  beamer_presentation: default
  ioslides_presentation:
    logo: hafro_logo.svg
    smaller: yes
    transition: faster
    widescreen: yes
  slidy_presentation: default
subtitle: Day 1
---

# Course outline

## Course outline
* The lecture series will tackle some recent development in R. The intended structure of the course is to have a weekly seminar/practical lab which introduces a particular topic followed by a small assignment as well as suggested further readings.

* The series is intended for those that have not very much experience in using R as their daily working environment. The instructors will be Bjarki Þór Elvarsson and Einar Hjörleifsson.

* The seminars will be in English, this being because the R environment is in that language as well as to accommodate potential participants whose native language may be other then Icelandic.

## List of potential topics and dates
<div class="columns-2">

* Friday, 22. January 2016 - 13:30 - 15:30
    - RStudio and project management
    - Basis of reproducible analysis and document writing
    - Some R base
    
* Friday, 19. January 2016 - 13:30 - 15:30
    - Getting data into R - introduction
    - Introduction to the grammar of graphics 
        + the ggplot2 library
    
* Friday, 5. February 2016 - 13:30 - 15:30
    - Getting data into R - beyond introduction 
    - Introduction to the grammar of data 
        + the dplyr library

* Friday, 12. January 2016 - 13:30 - 15:30
    - Data exploration based on grammar of data and graphics

* Friday, 19. February 2016 - 13:30 - 15:30
    - More data exploration based on grammar of data and graphics

</div>


## Today
* Introduction to the R-environment
* RStudio and project management
  - Using R
* Basis of reproducible analysis and document writing

```{r fig.width=6, fig.height=3,echo=FALSE}
library(png)
library(grid)
library(gridExtra)
img <- readPNG("Rlogo-1.png")
img2 <- readPNG("RStudio-Ball.png")

grid.arrange(rasterGrob(img),rasterGrob(img2),ncol=2)
```

## Why R? (.. or why not Excel)
R has become the lingua france of statistical analysis and data wrangling

* It’s free! If you are a teacher or a student, the benefits are obvious
* It runs on a variety of platforms including Windows, Unix and MacOS
* It provides an unparalleled platform for programming new statistical
methods in an easy and straightforward manner
* It offers powerful tools for data exploration and presentation


## The origins of R | The S language
<div class="columns-2">

```{r fig.width=6, fig.height=3,echo=FALSE}
library(png)
library(grid)
library(gridExtra)
img <- readPNG("bell.png")

grid.raster(img)
```

* The early days of statistical computing used fortran (and punchcards)
* The S language conceived to allow interactive data analysis, combining fortran routines 
* Developed at Bell labs in the late 70's by John Chambers
* One of many languages developed at the time for interactive data analysis and calculations (SAS, SPSS, Matlab)
* Commercially implemented as Splus
</div>

## The origins of R | Development of R
<div class="columns-2">

```{r fig.width=3, fig.height=6,echo=FALSE}
library(jpeg)
library(grid)
library(gridExtra)
img <- readJPEG('Ihaka-Gentleman.jpg')
grid.raster(img)
```

* The development of R started in the early 90's by Ross Ihaka and Robert Gentleman in New Zealand
* The main driver behind the development was to build a teaching tool for statistics based loosely on the S language
* It quickly became very popular among statistician and has slowly been seeping into other communities 

</div>

# What is R 

## What is R
* Program that does statistics
* A programming language
* Data crunching environment 

# RStudio

## Interacting with R


## RStudio
* RStudio allows the user to run R in a more user-friendly environment.  It is open- source (i.e. free) and available at \url{www.rstudio.com}
* built to help you write R code, run R code, and analyze data with R
* text editor, version control, project handling, markdown support, keyboard shortcuts, debugging tools, and much more

